A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)

### 2019-12-07 Simulation/Writeup/NoteConvergence.Rnw
fiest paragraph of theoretical background

The convergence properties of the MICE algorithm are investigated by means of model-based simulation in \proglang{R} \citep{R}. The results of this simulation study are guidelines for assessing convergence of this MI method. These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}. 
%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

@@

The method of multiple imputation The properties of multiple imputation procedures have been studied

@@

A fundamental assumption of MICE is convergence of the algorithm. MICE is a type of Markov chain Monte Carlo (MCMC) algorithm. In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence. Hence we use convergence diagnostics to flag non-convergence. But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.

@@

Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07} \textbf{[Expand or remove Rhat here! Refer to future section?]}. Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} \textbf{[explain why this is appropriate, and numerical is not]}. Visually assessing convergence is not only difficult to the untrained eye, it might also be futile. The convergence properties of MI algorithms lack scientific consensus \citep{taka17}, and some default MICE techniques might not converge to stable distributions at all \citep{murr18}. Moreover, convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.

>>>>

The convergence properties of MICE are investigated by means of model-based simulation in \proglang{R} \citep{R}. The results of this simulation study are guidelines for assessing convergence of the MI algorithm. These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}. 
%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

@@

The methodology for diagnosing convergence of multiple imputation algorithms has not been systematically studied \cite{buur18}. Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, XYZ, XYZ). This section provides a brief review of how convergence is defined in the context of iterative algorithmic procedures, what diagnostic tools are available, and which of these may be applicable to MI algorithms.

@@

The convergence properties of multiple imputation algorithms lack scientific consensus \citep{taka17}. A review by \citep{murr18}  and some default MICE techniques might not converge to stable distributions at all \citep{murr18}. Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.  Moreover, 
A fundamental assumption of MICE is convergence of the algorithm. MICE is a type of Markov chain Monte Carlo (MCMC) algorithm. In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence. Hence we use convergence diagnostics to flag non-convergence. But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.

@@

Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07} \textbf{[Expand or remove Rhat here! Refer to future section?]}. Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} \textbf{[explain why this is appropriate, and numerical is not]}. Visually assessing convergence is not only difficult to the untrained eye, it might also be futile.
